Efficient Computation of Extended Sources

### 2021-06-07 repro/papers/sseprecond/oldstuff.tex
initial commit - pruned stuff

>>>>

\section{Discussion}

The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}. In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $\bff$ in \ref{eqn:esis} is treated as a function of $\bf{c}$, which is in turn minimized. Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $\bff$ at every step of an iterative method for the outer problem of minimization over $\bf{c}$. Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.

This paper restricts the choice of wave physics embodied in
$L[\bf{c}]$ to linear acoustics.  Thus $\bf{c}=(\kappa,\rho)$ is the
pair (bulk modulus, density) of positive functions of spatial
position. The principal characteristics assumed of sources $\bff=$ are localization and isotropy. That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
\begin{equation}
\label{eqn:ptsrc}
f_{\rm pt}(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
\end{equation}
Source locations $\bx_s$ are assumed known (as is natural for active source methods), and the source {\em wavelet} $w(t)$ is to be found as part of the solution of \ref{eqn:esi}. A convenient choice of penalty operator $A$ is then multiplication by the {\em offset} $|\bx-\bx_s|$: its null space consists precisely of distributions of the form \ref{eqn:ptsrc}.

More complex radiation characteristics may be accommodated with similar but more complex choice for the operator $A$. Since the core analysis is similar, this paper deals only with the simplest case \ref{eqn:ptsrc}.

The wavefield $u$ in this work is presumed to exist in all of Euclicean space-time: in other words, boundary effects have been removed from $u$, by surface-related multiple elimination or other techniques. Predicted data traces are simply the samples of one or more components of $u$ at receiver points $\bx_r$.

%The domain of wave propagation is denoted $\Omega$, and two subsets of its boundary are identified: $\Gamma_s$ contains the support (non-zeroes) of the (extended) sources and in particular the locations $\bx_s$ of physical sources. Data trace locations $\bx_r$ lie in another subset $\Gamma_r$. The data projection operator $P$ samples the acoustic field $u$ on $\Gamma_r$, and includes a finite aperture mute.

Achieving a small residual in the source subproblem \ref{eqn:esis} for arbitrary $c$ is a critical part of the convergence mechanism for the extended source approach to FWI \cite[]{HuangNammourSymesDollizal:SEG19}. However, for arbitrary data $d$, coefficient array $c$, and source location $\bx_s$, there does not exist an isotropic point source $f_{\rm pt}$ of the form \ref{eqn:ptsrc} for which $PL[c]^{-1}f_{\rm pt} \approx d$. This is so even if there does exist a coefficient array $c^*$ and point source $f_{\rm pt}^*$ located at $\bx_s$ with wavelet $w^*(t)$ for which $d = PL[c^*]^{-1}f_{\rm pt}^*$, if $c$ differs substantially from $c^*$. It is this fact that underlies the effectiveness of the extended source approach to FWI:  driving the residual in the problem \ref{eqn:esi} towards zero necessarily requires modifying $c$ to resemble $c^*$, assuming the latter exists, as well as pushing the optimal $f$ towards the null space of $A$. Conversely, the source subproblem \ref{eqn:esis} does not have a small residual solution of point source form \ref{eqn:ptsrc}, in general, for a given choice of $c$, even if that were the case for a different choice ($c^*$). Therefore a more general class of source models than that specified by the condition \ref{eqn:ptsrc} must be admitted to the feasible set for the source subproblem \ref{eqn:esis} if it is to serve as the inner problem in a variable projection formulation of extended source FWI.

The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not {\em a priori} required to have point support. These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$. This observation is a by-product of the approximate solution for \ref{eqn:esis} constructed below. The construction of the approximate solution is closely related to the {\em time reversal} method of photoacoustic tomography (see \cite{StefanovUhlmannIP:09} and references cited there).

The next section gives a precise description of an acoustic version of problem \ref{eqn:esis} for an idealized {\em crosswell} configuration, in which source located on a surface $z=z_s$ generate waves that propagate to to a receiver surface $z=z_r > z_s$. In principle, {\em diving waves} can be described with similar mathematics, but with some additional complications, so I do not treat the diving wave case explicitly in this paper. The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}. There are no mathematical results at present concerning reflected wave inverse problems treated by source extension methods, though there are tantalizing numerical clues \cite[]{LeeuwenHerrmannWRI:13,Warner:14,Warner:16,LeeuwenHerrmann:16,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b}.

The following section describes the construction of the approximate
inverse and explains the conditions under which accuracy should be expected. I observe that the approximate inverse is approximately the adjoint of the modeling operator $PL[c]^{-1}$ with respect to weighted norms in its domain and range (see \cite{HouSymes:15} for a similar observation in a different context). This fact immediately suggests use of the approximate inverse as a preconditioner to accelerate convergence of Krylov space methods such as conjugate gradient iteration applied to the least squares problem \ref{eqn:esis}. However, its straightforward application in the preconditioned conjugate gradient (PCG) method \cite[]{Golub:2012} involves explicitly a version the so called Dirichlet-to-Neumann (D2N) operator, which is computationally awkward. I show how to reorganize the PCG iteration so that only solutions of $L[c]u=f$ are required. The penultimate section presents some 2D examples in the crosswell configuration, illustrating the accuracy of the approximate inversion and the accelerated convergence of PCG. The examples are chosen to emphasize that surface extended sources can be constructed to fit more or less arbitrary data, unlike point sources - as mentioned before, this approximate invertibility property of the modeling operator is of critical importance in the application to extended FWI. The paper ends with a discussion of several important matters not addressed here, and a restatement of the conclusions.

The intent of the main body of this paper is to present the formal structure of the extended source subproblem and the way in which this structure leads to accelerated numerical solution. This formal structure is however supported by a rigorous foundation. In order to avoid disrupting the formal account, this foundation material is relegated to an appendix.

\section{Prototype Numerical Examples}
I present a collection of simple examples that illustrate the features of the surface source extension claimed in preceding sections.

I used the IWAVE acoustic staggered grid package to carry out these calculations. This package implements (2,2k) schemes for k=1,2,..., and outputs traces (of either velocity or pressure at any point in space via multilinear interpolation. The discretized modeling operator is thus of second order accuracy, though as usual I have used higher order in space to reduce grid dispersion.

Source injection is implemented as the adjoint of trace sampling, resulting in another second-order error [REFERENCES].

The data is a single shot gather, with a source at coordinates $x_s=z_s=3000$ m (units of length are meters in all cases). The receiver line occupies $1500 \le x_r \le 5500$ m, with receiver depth $z_r=1000$ m.  Extended sources occupy $1500 \le x_s \le 5500$ m, with the same depth $z_s=3000$ m as the ``physical'' source used to generate the data. This region turns out to be adequate to represent the extended sources that approximately invert the data, for the cases examined below. An algorithm to automatically identify an appropriate region can be based on the ideas developed by \cite{Fu:Geo17}.

I have used absorbing boundary conditions (split-field PML) on all four sides of the 4000 m (vertical) $\times$ 8000 m (horizontal) simulation domain. Evidently inclusion of a free surface is important to the application of the ideas explained here to diving wave marine data, and I have not addressed the necessary modifications here.

The Dirichlet-to-Neumann operator $\Lambda$ is an essential part of the inner problem preconditioner just presented. Implementation can be accomplished in several ways:
\begin{itemize}
\item \cite{tenKroode:12} suggests using a one-way operator;
\item if both pressure and normal partical velocity are measured (or simulated), then the two are related by $\Lambda$ and the velocity component can simply be used as the output;
\item presence of a free surface implies all of the usual problems, such as the need for removal of receiver-side ghosts. On the other hand, if the free surface is within a quarter-wavelength throughout the useful bandwidth of the data, then the ghosted data differs from $\Lambda p$ by a time integration and a scale factor, a fact used to good effect by \cite{HouSymes:15}.
\end{itemize}

In the examples, I have used the second observation. With a finite difference implementation of the pressure-velocity system \ref{eqn:awe}, velocity components are available ``for free'', short-circuiting explicit computation of $\Lambda$.

The  {\tt project/SConstruct} script is set up to carry out the necessary computations on grids with spacings $\Delta x = \Delta z = $ 20, 10, and 5 m, with a jump of roughly 8 in computation time resulting from each refinement. For present purposes, the coarsest (20 m) grid seems to be sufficient, and that is the grid used in the examples presented below. The source pulses are chosen so that the computation is reasonably accurate. For the 20 m grid case, I use a zero-phase trapezoidal bandpass filter source with corner frequencies of 1.0, 2.0, 7.5, and 12.5 Hz.

The IWAVE asg driver has been set up to recognize the case {\tt deriv=0} as defining the map from source (right-hand side in the pressure equation) to data (pressure) traces. The adjoint to this map, as explained above, is reverse-time propagation of the data traces as pressure sources, followed by scaling (formula \ref{eqn:sadj}). The approximate inverse is computed by application of the Dirichlet-to-Neumann operator to the pressure traces to produce corresponding velocity traces, followed by injection as pressure sources and reverse time propagation, followed by another application of the Dirichlet-to-Neumann map and scaling (formulas \ref{eqn:appinv}, \ref{eqn:adj}).

The script implements these operatations step-by-step via calls to IWAVE, Madagascar, and SU commands. A peculiarity of the {\tt asg.x} driver needs to be mentioned: it is based on an un-scaled version of the constitutive law defect source representation, that is, {\tt asg.x} approximately computes the solution of the system \ref{eqn:awedata} with the first equation replaced by
\begin{equation}
\label{eqn:asgdata}
\frac{\partial p}{\partial t}  =  - \kappa \nabla \cdot \bv +
h \delta(z-z_s).
\end{equation}
Denote by $S_{\rm asg}[c]$ the forward map produced by {\tt asg.x}. Then comparison of \ref{eqn:awedata} and \ref{eqn:asgdata} reveals that
\begin{equation}
\label{eqn:sreln}
S[c]=S_{\rm asg}[c]\kappa
\end{equation}
where $\kappa$ is shorthand for the operator of multiplication by $\kappa$. Accordingly, and approximate inverse for $S_{\rm asg}[c]$ is
\[
I \approx S_{\rm asg}[c]^{\dagger}S_{\rm asg}[c] = S_{\rm asg}[c]^{\dagger}S[c]\kappa^{-1}
\]
Since $S[c]S[c]^{\dagger} \approx I \approx S[c]^{\dagger}S[c]$, it follows that
\[
S_{\rm asg}[c]^{\dagger}=\kappa S[c]^{\dagger} = \kappa\Lambda[c]_s S[c]^T \Lambda[c]_r 
\]
\begin{equation}
\label{eqn:asginv}
= \kappa \Lambda[c]_s \kappa S_{\rm asg}[c]^T\Lambda[c]_r
\end{equation}
This is the approximate inverse computed in the examples. Note that
only the values of $\kappa$ near the source datum $z=z_s$ play a role
in the relation \ref{eqn:sreln} or in the definition \ref{eqn:asginv}
of the IWAVE ASG approximate inverse.
